Automated Program Verification and Synthesis of Embedded \& Cyber-physical Systems: Advances and Challenges

### 2017-08-08 paper.tex
final additions

\textbf{(RP1)} Cordeiro, Fischer, and Marques-Silva proposed the first SMT-based BMC for full C programs, called Efficient SMT-Based Context-Bounded Model Checker (ESBMC)~\cite{Cordeiro12}, which was later extended to support C++98 programs~\cite{ECBS13}, CUDA programs~\cite{cudalucas,Pereira17}, and Qt-based consumer electronics applications~\cite{Sousa15}. This approach was also able to find undiscovered bugs related to arithmetic overflow, buffer overflow, and invalid pointer, in standard benchmarks, which were later confirmed by the benchmarks' creators ({\it e.g.}, NOKIA, NEC, NXP, and VERISEC)~\cite{CordeiroF11,Cordeiro12}. Other SMT-based BMC approaches have also been proposed and implemented~\cite{MerzFS12}, but the coverage and verification time of all existing ones are still limited to specific program classes, especially for those that contain intensive floating-point arithmetic and dynamic memory allocation~\cite{Beyer14,BeyerSVCOMP15}. One possible research direction is to bridge the gap between BMC tools and SMT solvers, propose background theories, and develop more efficient decision procedures, in order to handle specific program classes. In this particular research direction, the European Project SC$^2$ (Satisfiability Checking and Symbolic Computation: uniting two communities to solve real problems) aims to further extend background theories supported by SMT solvers, in order to solve problems related to security and safety of computer systems (or large mathematical problems) by the development of radically improved software tools~\cite{DBLP:journals/cca/AbrahamA0BBBCDE16}.

\textbf{(RP2)} The SMT-based BMC approach proposed by Cordeiro, Fischer, and Marques-Silva was further developed to verify correct lock acquisition ordering and the absence of deadlocks, data races, and atomicity violations in multi-threaded software based on POSIX and CUDA libraries~\cite{CordeiroF11,Pereira17}, considering monotonic partial-order reduction~\cite{KahlonWG09} and state-hashing techniques, in order to prune state-space exploration~\cite{morse15}. Recent advances for verifying multi-threaded C programs have been proposed to speed up the verification time, which significantly prune the state-space exploration~\cite{Inverso14,civl15}, and also extensions to verify the Debian GNU/Linux distribution based on a context- and thread-sensitive abstract interpretation framework~\cite{DBLP:conf/kbse/KroeningPSW16}; however, the classes of concurrent programs ({\it e.g.}, CUDA, OpenCL, and MPI) that can be verified is still very limited. One possible research direction is to further extend BMC of multi-threaded C programs via Lazy Sequentialization~\cite{Inverso14}, in order to analyze unsatisfiability cores~\cite{Grumberg05}, with the goal to remove redundant behaviour or analyze interpolants~\cite{McMillan11} to prove non-interference of context switches.

\textbf{(RP3)} Novel approaches to model check embedded software using \textit{k}-induction and invariants were proposed and evaluated in the literature, which demonstrate its effectiveness in some real-life embedded-system applications~\cite{Gadelha15,Brain15,Rocha15,Donaldson10,Rocha17}; however, the main challenge remains still open, {\it i.e.}, to compute and strengthen loop invariants to prove program correctness and timeliness in a more efficient and effective way, in order to be competitive with other model-checking approaches. In particular, invariant-generation algorithms have substantially evolved over the last years, with the goal of discovering inductive invariants of programs~\cite{pips:2013,Henry:2012} or continuously refine them during verification~\cite{Beyer15}; however, there is still a lack of studies for exploiting the combination of different invariant-generation algorithms ({\it e.g.}, interval analysis, linear inequalities, polynomial equalities and inequalities) and how to strengthen them during verification, in order to ensure system robustness w.r.t. implementation aspects. Finally, optimal configurations for invariant generators are still of paramount importance, given that applications with code based on different aspects may benefit from different setups.

@@

{\textbf{(RP5)} Although ESBMC \cite{Cordeiro12} was extended to support C/C++ and some variants, new application interfaces and programming languages are often developed, which require suitable software verification tools. Indeed, it would be interesting if a new programming language model could be loaded, which along with a BMC core could check different programs. Some work towards that was already presented by Sousa, Cordeiro, and Filho~\cite{Garcia16}, which employed operational models for checking Qt-based programs from consumer electronics. In summary, the BMC core (in that case, ESBMC) is not changed, but instead an operational model, which implements the behavior and features of Qt libraries, is used to provide the new code structure to be checked. This approach could adopt the clang compiler~\cite{Lopes:2014:GSL:2692607}, which is already widely used in industry~\cite{clang-usage}, in order to produce an Abstract Syntax Tree (AST) for a given programming language ({\it e.g.}, C/C++/ObjectiveC/ObjectiveC++). This AST can then be converted either to an intermediate representation (IR) of the underlying verifier or to LLVM bitcode that is already produced by clang. Such research problem is closely related to the first one \textbf{(RP1)} and has the potential to devise a new paradigm in software verification.

@@

\textbf{(RP7)} Embedded systems are typically implemented in low- and medium-level programming languages ({\it e.g.}, ANSI-C/C++/Java). Given the availability of different programming languages for writing code for ECPS, the provision of a unified framework able to tackle low- and high-level properties, as well as system behavior, is needed to achieve code coverage in real applications. The clang compiler~\cite{Lopes:2014:GSL:2692607} seems to offer a simple and strong approach to lead software verification to another degree of completeness so that every implementation aspect of an embedded system can be addressed and evaluated. In particular, by using the clang compiler as front-end for a software verification tool, we avoid the need for maintaining a proprietary front-end (a real challenge nowadays, given that the C and C++ standards are rapidly evolving), and then focus on the main tool objective, the formal program verification. Here, we provide only operation models to tackle low- and high-level system properties via an abstract representation of the associated libraries, which conservatively approximates their semantics. The clang compiler offers an industrial quality analyzer and an application programming interface to access and traverse its internal AST, which can then be used by software verifiers to generate their IR or alternatively analyze the LLVM bitcode produced by clang.

@@

As already mentioned in section \ref{sec:model}, problems that are not directly related to code correctness may be tackled with formal methods, as long as they are conveniently modeled. Ara\'ujo {\it et al.} \cite{Araujo16} employed SMT for handling non-convex optimization problems, with the goal of ensuring optimal solutions. Indeed, it is possible to recursively re-constrains a model checking procedure, by using the current counterexample as a seed for the next run, until a global minimum is achieved. Nonetheless, the elapsed execution time is still a bottleneck for some applications, such as changes in UAV trajectories, but that can be improved, through the use of simplification techniques and problem constraints.

In the same fashion, other problems may benefit from the recursive-refinement approach provided by SMT-based verification schemes. For instance, de Jesus {\it et. al} \cite{antennasbrt} developed a simplified methodology for antenna alignment that makes use of many modules already available in satellite TV receivers, along with a suitable structure for performing position correction. In summary, a reference signal is evaluated, until an (approximately) error-free reception is acquired, which can be measured through a simple transport stream (TS) error-indication interface and further refined with other parameters, such as carrier-to-noise ratio and signal level. This way, by using an SMT-based verification method, one may map the TS error indication and other performance figures as properties and then recursively perform realignment procedures, until a counterexample is not output. It is worth noticing that the timing requirements are more relaxed than what was mentioned in the work developed by Ara\'ujo {\it et al.}~\cite{Araujo16}, which can be seen as an advantage for such an implementation.

Regarding behavior correctness, many other applications can be tackled, such as the ones related to automatic classification. Amoedo {\it et al.} \cite{amoedo} proposed a modulation classification methodology, with the goal of spectrum sensing for opportunistic transmissions. Besides a correct software implementation, {\it i.e.}, source code, which is already covered by formal methods already available in literature \cite{esbmc_filter,Cordeiro12,Clarke04}, one may argue if the proposed methodology is able to recognize modulations under some reception condition, such as severe analog interference. This way, it would be interesting to develop a verification module for transmission and reception software, which could be capable of creating inputs related to sampled radio frequency signals under a variety of conditions, unifying source code and behavior verification. In addition, that kind of implementation usually employs digital filters with specific parameters, which could also be tackled.

>>>>

\textbf{(RP1)} Cordeiro, Fischer, and Marques-Silva proposed the first SMT-based BMC for full C programs, called Efficient SMT-Based Context-Bounded Model Checker (ESBMC)~\cite{Cordeiro12}, which was later extended to support C++98 programs~\cite{ECBS13}, CUDA programs~\cite{cudalucas,Pereira17}, and Qt-based consumer electronics applications~\cite{Sousa15}. This approach was also able to find undiscovered bugs related to arithmetic overflow, buffer overflow, and invalid pointer, in standard benchmarks, which were later confirmed by the benchmarks' creators ({\it e.g.}, NOKIA, NEC, NXP, and VERISEC)~\cite{CordeiroF11,Cordeiro12}. Other SMT-based BMC approaches have also been proposed and implemented~\cite{MerzFS12}, but the coverage and verification time of all existing ones are still limited to specific program classes, especially for those that contain intensive floating-point arithmetic and dynamic memory allocation~\cite{Beyer14,BeyerSVCOMP15}. One possible research direction is to bridge the gap between BMC tools and SMT solvers, propose background theories, and develop more efficient decision procedures, in order to handle specific program classes. In this particular research direction, the European Project SC$^2$ (Satisfiability Checking and Symbolic Computation: uniting two communities to solve real problems) aims to further extend background theories supported by SMT solvers, in order to solve problems related to security and safety of computer systems (or large mathematical problems), through the development of radically improved software tools~\cite{DBLP:journals/cca/AbrahamA0BBBCDE16}.

\textbf{(RP2)} The SMT-based BMC approach proposed by Cordeiro, Fischer, and Marques-Silva was further developed to verify correct lock acquisition ordering and absence of deadlocks, data races, and atomicity violations in multi-threaded software based on POSIX and CUDA libraries~\cite{CordeiroF11,Pereira17}, considering monotonic partial-order reduction~\cite{KahlonWG09} and state-hashing techniques, in order to prune state-space exploration~\cite{morse15}. Recent advances for verifying multi-threaded C programs have been proposed to speed up verification times, which significantly prune state-space exploration~\cite{Inverso14,civl15}, and also extensions to verify the Debian GNU/Linux distribution, based on a context- and thread-sensitive abstract interpretation framework~\cite{DBLP:conf/kbse/KroeningPSW16}, were made available; however, the set of concurrent-program classes ({\it e.g.}, CUDA, OpenCL, and MPI) that can be verified is still very limited. One possible research direction is to further extend BMC of multi-threaded C programs via Lazy Sequentialization~\cite{Inverso14}, in order to analyze unsatisfiability cores~\cite{Grumberg05}, with the goal of removing redundant behaviour or analyzing interpolants~\cite{McMillan11} and then prove non-interference of context switches.

\textbf{(RP3)} Novel approaches to model check embedded software using \textit{k}-induction and invariants were proposed and evaluated in literature, which demonstrate its effectiveness in some real-life embedded-system applications~\cite{Gadelha15,Brain15,Rocha15,Donaldson10,Rocha17}; however, the main challenge remains still open, {\it i.e.}, to compute and strengthen loop invariants for proving program correctness and timeliness in a more efficient and effective way, in order to be competitive with other model-checking approaches. In particular, invariant-generation algorithms have substantially evolved over the last years, with the goal of discovering inductive invariants of programs~\cite{pips:2013,Henry:2012} or continuously refine them during verification~\cite{Beyer15}; however, there is still a lack of studies for exploiting the combination of different invariant-generation algorithms ({\it e.g.}, interval analysis, linear inequalities, polynomial equalities and inequalities) and how to strengthen them during verification, in order to ensure system robustness w.r.t. implementation aspects. Finally, optimal configurations for invariant generators are still of paramount importance, given that applications with code based on different aspects may benefit from different setups.

@@

{\textbf{(RP5)} Although ESBMC \cite{Cordeiro12} was extended to support C/C++ and some variants, new application interfaces and programming languages are often developed, which require suitable software verification tools. Indeed, it would be interesting if a new programming language model could be loaded, which along with a BMC core could check different programs. Some work towards that was already presented by Sousa, Cordeiro, and Filho~\cite{Garcia16}, which employed operational models for checking Qt-based programs from consumer electronics. In summary, the BMC core (in that case, ESBMC) is not changed (it is still C/C++ code), but instead an operational model, which implements the behavior and features of Qt libraries, is used to provide the new code structure to be checked. This approach could adopt the clang compiler~\cite{Lopes:2014:GSL:2692607}, which is already widely used in industry~\cite{clang-usage}, in order to produce an Abstract Syntax Tree (AST) for a given programming language ({\it e.g.}, C/C++/ObjectiveC/ObjectiveC++). This AST can then be converted either to an intermediate representation (IR) of the underlying verifier or to LLVM bitcode that is already produced by clang. Obviously, a completely new language may present new structures and resources; however, if different inputs are parsed and converted into a form that preserves every aspect of their structure, that can then be analyzed and verified in an unified way. Such research problem is closely related to the first one \textbf{(RP1)} and has the potential to devise a new paradigm in software verification.

@@

\textbf{(RP7)} Embedded systems are typically implemented in low- and medium-level programming languages ({\it e.g.}, ANSI-C/C++/Java). Given the availability of different programming languages for writing code for ECPS, the provision of a unified framework able to tackle low- and high-level properties, as well as system behavior, is needed to achieve code coverage in real applications. The clang compiler~\cite{Lopes:2014:GSL:2692607} seems to offer a simple and strong approach to lead software verification to another degree of completeness so that every implementation aspect of an embedded system can be addressed and evaluated. In particular, by using the clang compiler as front-end for a software verification tool, we avoid the need for maintaining a proprietary front-end (a real challenge nowadays, given that the C and C++ standards are rapidly evolving), and then focus on the main objective: formal program verification. Here, we provide only operation models to tackle low- and high-level system properties via an abstract representation of the associated libraries, which conservatively approximates their semantics. The clang compiler offers an industrial quality analyzer and an application programming interface to access and traverse its internal AST, which can then be used by software verifiers to generate their IR or alternatively analyze the LLVM bitcode produced by clang.

@@

As already mentioned in section \ref{sec:model}, problems that are not directly related to code correctness may be tackled with formal methods, as long as they are conveniently modeled. Ara\'ujo {\it et al.} \cite{Araujo16} employed SMT for handling non-convex optimization problems, with the goal of ensuring optimal solutions. Indeed, it is possible to recursively re-constrains a model checking procedure, by using the current counterexample as a seed for the next run, until a global minimum is achieved. Nonetheless, elapsed execution times are still a bottleneck for some applications, such as changes in UAV trajectories, but that can be improved, through the use of simplification techniques and problem constraints.

In the same fashion, other problems may benefit from the recursive-refinement approach provided by SMT-based verification schemes. For instance, de Jesus {\it et. al} \cite{antennasbrt} developed a simplified methodology for antenna alignment that makes use of many modules already available in satellite TV receivers, along with a suitable structure for performing position correction. In summary, a reference signal is evaluated, until an (approximately) error-free reception is acquired, which can be measured through a simple transport stream (TS) error-indication interface and further refined with other parameters, such as carrier-to-noise ratio and signal level. This way, by using an SMT-based verification method, one may map the TS error indication and other performance figures as properties and then recursively perform realignment procedures, until a counterexample is not output. It is worth noticing that associated timing requirements are more relaxed than what was mentioned in the work developed by Ara\'ujo {\it et al.}~\cite{Araujo16}, which can be seen as an advantage for such an implementation.

Regarding behavior correctness, many other applications can be tackled, such as the ones related to automatic classification. Amoedo {\it et al.} \cite{amoedo} proposed a modulation classification methodology, with the goal of spectrum sensing for opportunistic transmissions. Besides a correct software implementation, {\it i.e.}, source code, which is already covered by formal methods already available in literature \cite{esbmc_filter,Cordeiro12,Clarke04}, one may argue if the proposed methodology is able to recognize modulations under some reception condition, such as severe analog interference. This way, it would be interesting to develop a verification module for transmission and reception software, which could be capable of creating inputs related to sampled radio frequency signals, under a variety of conditions, and then unifying source code and behavior verification. In addition, that kind of implementation usually employs digital filters with specific parameters, which could also be tackled. Indeed, communication aspects of ECPSs could be tackled, as a module in system verification procedures.
